---
format: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, cache = T)
```

This script outlines the analysis and code used in James et al. 2024 to analyze the effect of habitat loss on ecosystem-scale trophic function. Percent cover data (FHAP_pc.csv) from the Florida Fish and Wildlife Conservation Commission Fisheries Habitat Assessment Program were used to generate habitat maps and combined with basal resource use from stable isotope mixing models (mm3spp.csv) from [James et al. 2022](https://doi.org/10.1093/icesjms/fsac112) to generate *E*-scapes for bay anchovy, pinfish, and pink shrimp from 2007-2019. This can also be found as an [R script](sgE-scape_analysis.R). 


## Generate habitat maps from monitoring data
Benthic habitat maps were generated based on spatial interpolation of benthic cover data from the Florida Fish and Wildlife Commission Fish Habitat Assessment Program. We used ordinary kriging with the `gstat` package to create a habitat map for seagrass, algae, and total leaf area for each basin in each year from 2007 to 2019 (with the exception of 2013 due to not having data on leaf area) at a 10 x 10 m cell size. For each habitat map, the variogram was automatically fit with the `autofitVariogram()` function in the `automap` package. To create a habitat map for mangrove, a 500 m width buffer was placed around the mangrove islands that bordered each basin. A 10 x 10 m raster was created for each basin, and cells were assigned as mangrove (value of 1) if they fell within the 500 m buffer and given a value of 0 if outside of the buffer. 
```{r, eval=F}

# load libraries 
library(tidyverse)
library(sf)
library(stars)
library(gstat)
library(automap)

# load shape files for each basin and make empty grid
# Rankin shape file
ran = st_read('gis/Rankin.shp') %>% st_transform(32617)

# make empty stars raster with 10 m cell size
ran_grid = st_bbox(ran) %>%
  st_as_stars(dx = 10) %>%
  st_set_crs(32617) %>%
  st_crop(ran)

# Johnson shape file
jon = st_read('gis/Johnson.shp') %>% st_transform(32617)

# make empty stars raster with 10 m cell size
jon_grid = st_bbox(jon) %>%
  st_as_stars(dx = 10) %>%
  st_set_crs(32617) %>%
  st_crop(jon)

# Whipray shape file
whp = st_read('gis/Whipray.shp') %>% st_transform(32617)

# make empty stars raster with 10 m cell size
whp_grid = st_bbox(whp) %>%
  st_as_stars(dx = 10) %>%
  st_set_crs(32617) %>%
  st_crop(whp)

# Create mangrove habitat map for each basin 
man = st_read('gis/FlBayMan500mBuf.shp')

# Rankin 
man_ran =  man |> st_intersection(ran) |> 
  st_rasterize(ran_grid) %>% 
  mutate(Acres = if_else(Acres > 0, 1, 0))
# Johnson
man_jon =  man |> st_intersection(jon) |>
  st_rasterize(jon_grid) %>% 
  mutate(Acres = if_else(Acres > 0, 1, 0))

# Whipray
man_whp =  man |> st_intersection(whp) |>
  st_rasterize(whp_grid) %>% 
  mutate(Acres = if_else(Acres > 0, 1, 0))

# combine into tibble
df_man = tibble(BASIN = c('RAN', 'JON', 'WHP'),
                man_r = c(list(man_ran), list(man_jon), list(man_whp)))

# Seagrass, algae, and total leaf area
df = st_read("FHAP_pc.csv",
             options=c("X_POSSIBLE_NAMES=LON","Y_POSSIBLE_NAMES=LAT")) |> 
  st_set_crs(4326) %>% st_transform(32617) |> 
  group_by(BASIN, YEAR) |> 
  nest() |> 
  mutate(sg = map(data, \(data) 
                  autofitVariogram(data$SG ~ 1, as(data, "Spatial"))),
         alg = map(data, \(data) 
                   autofitVariogram(data$ALG ~ 1, as(data, "Spatial"))),
         lab = map(data, \(data) 
                   autofitVariogram(data$LAB ~ 1, as(data, "Spatial"))),
         sg_r = case_when(
           BASIN == 'RAN' ~ map2(data,sg, \(data,sg)
                                 krige(SG~1, locations = data,
                                       newdata = ran_grid, model = sg$var_model)),
           BASIN == 'JON' ~ map2(data,sg, \(data,sg)
                                 krige(SG~1, locations = data,
                                       newdata = jon_grid, model = sg$var_model)),
           BASIN == 'WHP' ~ map2(data,sg, \(data,sg)
                                 krige(SG~1, locations = data,
                                       newdata = whp_grid, model = sg$var_model))),
         alg_r = case_when(
           BASIN == 'RAN' ~ map2(data,alg, \(data,alg)
                                 krige(ALG~1, locations = data,
                                       newdata = ran_grid, model = alg$var_model)),
           BASIN == 'JON' ~ map2(data,alg, \(data,alg)
                                 krige(ALG~1, locations = data,
                                       newdata = jon_grid, model = alg$var_model)),
           BASIN == 'WHP' ~ map2(data,alg, \(data,alg)
                                 krige(ALG~1, locations = data,
                                       newdata = whp_grid, model = alg$var_model))),
         lab_r = case_when(
           BASIN == 'RAN' ~ map2(data,lab, \(data,lab)
                                 krige(LAB~1, locations = data,
                                       newdata = ran_grid, model = lab$var_model)),
           BASIN == 'JON' ~ map2(data,lab, \(data,lab)
                                 krige(LAB~1, locations = data,
                                       newdata = jon_grid, model = lab$var_model)),
           BASIN == 'WHP' ~ map2(data,lab, \(data,lab)
                                 krige(LAB~1, locations = data,
                                       newdata = whp_grid, model = lab$var_model))))

# create dataset of rasters for each year
df_r = df |>
  select(YEAR, BASIN, sg_r, alg_r, lab_r) |> 
  left_join(df_man, by = 'BASIN')


```
![Example kriged habitat maps (right column) from monitoring points (left column) for a) seagrass, b) total leaf area, c) algae, and d) mangrove. Left column map for d) mangrove is the habitat map used to generate the 500 m buffer. Habitat maps shown here were generated from 2019 data.](publication/FigS1.tiff) 

## Calculate Index of Energetic Importance (IEI) values
Combining the basal resource use and habitat maps, we calculated an index of energetic importance (IEI) for each of the four resource/habitat type combinations for each consumer (James et al. 2022a). Each IEI was calculated with the following formula: 
$$IEI_i=  \frac{f_{source_i}}{f_{habitat_i}} $$
where $f_{source_i}$ is the fraction of the contribution of source $i$ to the total source use of the consumer based on the results of the mixing model and $f_{habitat_i}$ is the fraction of habitat type $i$ that produces source $i$ to the overall area at a scale relevant to the movement range of the consumer (i.e., area of the circle around the sampling point). An example of resource/habitat type combination is amount of seagrass-derived production and the cover area of seagrass habitat. An IEI around one means that the consumer is using a resource $f_{source_i}$ around the same amount as the proportion of that resource’s habitat type $f_{habitat_i}$ relative to total area that the consumer forages. An IEI > 1 means that the consumer is using that source more than expected based on the proportion of that habitat type in the total foraging area, while the opposite is true for an IEI < 1. We calculated IEIs for each of the 3 consumer species by randomly generating 50 points within the kriged habitat maps for Johnson, Rankin, and Whipray basins (n = 150 for each species). The $f_{habitat_i}$ was calculated by taking the mean proportion cover of the habitat map within a 300 m radius buffer of the point with the `exactextractr` package. Some of the kriging models led to values for habitat cover that were < 0, and if this was the case, $f_{habitat_i}$ values < 0 were converted to 0 before calculating IEI values. IEI values were calculated using the habitat maps from 2019 to closely match when the sampling for basal resource use was conducted.

```{r, eval=F}
# load mixing model data
iso = read_csv('mm3spp.csv')

# random points within each basin
# set sample number 
n = 50

# set radius length
radius = 300

# filter raster to only 2019 to use for IEI values 
r19 = df_r |> 
  filter(YEAR == 2019)

# create buffer around random points for each basin
buf_ran = st_sample(ran, n) %>%
  st_as_sf %>%
  st_transform(32617)%>% # utm 17
  mutate(site = row_number(),
         BASIN = 'RAN')%>%
  st_buffer(dist = radius)

buf_jon = st_sample(jon, n) %>%
  st_as_sf %>%
  st_transform(32617)%>% # utm 17
  mutate(site = row_number(),
         BASIN = 'JON')%>%
  st_buffer( dist = radius)

buf_whp = st_sample(whp, n) %>%
  st_as_sf %>%
  st_transform(32617)%>% # utm 17
  mutate(site = row_number(),
         BASIN = 'WHP')%>%
  st_buffer( dist = radius)

# calculate f_habitat for each buffer 
ran_fhab = tibble(site = buf_ran$site,
                  BASIN = 'RAN') |> 
  mutate(f_sg = exact_extract(terra::rast(r19[r19$BASIN == 'RAN',]$sg_r[[1]]), buf_ran, 'mean', progress = F)[[1]],
         f_man = exact_extract(terra::rast(r19[r19$BASIN == 'RAN',]$man_r[[1]]), buf_ran, 'mean', progress = F)[[1]],
         f_epi = exact_extract(terra::rast(r19[r19$BASIN == 'RAN',]$lab_r[[1]]), buf_ran, 'mean', progress = F)[[1]],
         f_alg = exact_extract(terra::rast(r19[r19$BASIN == 'RAN',]$alg_r[[1]]), buf_ran, 'mean', progress = F)[[1]])


jon_fhab = tibble(site = buf_jon$site,
                  BASIN = 'JON') |> 
  mutate(f_sg = exact_extract(terra::rast(r19[r19$BASIN == 'JON',]$sg_r[[1]]), buf_jon, 'mean', progress = F)[[1]],
         f_man = exact_extract(terra::rast(r19[r19$BASIN == 'JON',]$man_r[[1]]), buf_jon, 'mean', progress = F)[[1]],
         f_epi = exact_extract(terra::rast(r19[r19$BASIN == 'JON',]$lab_r[[1]]), buf_jon, 'mean', progress = F)[[1]],
         f_alg = exact_extract(terra::rast(r19[r19$BASIN == 'JON',]$alg_r[[1]]), buf_jon, 'mean', progress = F)[[1]])


whp_fhab = tibble(site = buf_whp$site,
                  BASIN = 'WHP') |> 
  mutate(f_sg = exact_extract(terra::rast(r19[r19$BASIN == 'WHP',]$sg_r[[1]]), buf_whp, 'mean', progress = F)[[1]],
         f_man = exact_extract(terra::rast(r19[r19$BASIN == 'WHP',]$man_r[[1]]), buf_whp, 'mean', progress = F)[[1]],
         f_epi = exact_extract(terra::rast(r19[r19$BASIN == 'WHP',]$lab_r[[1]]), buf_whp, 'mean', progress = F)[[1]],
         f_alg = exact_extract(terra::rast(r19[r19$BASIN == 'WHP',]$alg_r[[1]]), buf_whp, 'mean', progress = F)[[1]])

# join together 
fhab = bind_rows(ran_fhab, jon_fhab, whp_fhab) |> 
  mutate(across(f_sg:f_alg, ~if_else(.x < 0, 0, .x)))

# calculate IEI for each species
IEI = fhab |> 
  slice(rep(1:n(), each = length(unique(iso$Species)))) |> 
  mutate(Species = rep(unique(iso$Species), times = nrow(fhab))) |> 
  left_join(iso, by = 'Species') |> 
  mutate(iei_sg = Seagrass/f_sg,
         iei_man = Mangrove/f_man,
         iei_epi = Epiphytes/f_epi,
         iei_alg = Algae/f_alg,
         iei_man = ifelse(iei_man == Inf, NA, iei_man)) |> 
  group_by(Species) |> 
  summarize(across(iei_sg:iei_alg, \(x) median(x, na.rm = T)))

```
![Boxplot of the index of energetic (IEI) values for each resource/habitat based on the 50 random points in each basin for each species calculated using the habitat maps for 2019. The horizontal black line represents IEI value of 1. IEI values > 1 means that the consumer is using that source more than expected based on the proportion of that habitat in the total foraging area, while the opposite is true for an IEI < 1.](publication/Fig2.tiff) 

## Make *E*-scape for each species, year, and basin
*E*-scapes were generated for each species at each basin between 2007 and 2019. IEI values were combined with habitat cover areas within a landscape foraging unit (grid cell with an area that corresponds to the movement range of the consumer) to calculate the habitat resource index (HRI). HRI was calculated with the following formula: 
$$HRI_x= \sum_{i=1}^{n}{\tilde{IEI_i}*f_{habitat_i}}$$
where $\tilde{IEI_i}$ is the median of the IEI for the species-specific source/habitat combination $i$ and $f_{habitat_i}$ is the fraction of habitat $i$ to the overall area within landscape foraging unit $x$. HRI is an index that represents a relative measurement of the quality of the habitats for producing the resources used by the consumer based on stable isotope analysis. An HRI value of 1 means that the area is producing the average amount of resources for the consumer. HRI values > 1 mean that the area is better for producing resources (i.e., more energetic resources) being used by the consumer, while HRI values < 1 shows that the habitats most important to the production of resources being used by the consumer are underrepresented within the landscape foraging unit. The $f_{habitat_i}$ was calculated by taking the mean value of the habitat map within each landscape foraging unit with the `exactextractr` package. Some of the kriging models led to values for habitat cover that were < 0, and if this was the case, $f_{habitat_i}$ values < 0 were converted to 0 before calculating HRI values. Each basin was subdivided into a grid of 600 m x 600 m landscape foraging units to generate an E-scape. This size was chosen to correspond to the 300 m radius (600 m diameter) buffer that was used to calculate the IEIs. Given the singular temporal point of resource use data, IEI values calculated for 2019 were applied across the entire time series.

```{r, eval=F}
library(exactextractr)
# set radius length for size of landscape foraging units
radius = 300

# make a grid of landscape foraging units
ran_grd = st_make_grid(ran, cellsize = radius*2) %>% 
  st_sf(geometry = .) %>%
  filter(lengths(st_intersects(., ran)) > 0) |> 
  mutate(BASIN = 'RAN')

jon_grd = st_make_grid(jon, cellsize = radius*2) %>% 
  st_sf(geometry = .) %>%
  filter(lengths(st_intersects(., jon)) > 0) |> 
  mutate(BASIN = 'JON')

whp_grd = st_make_grid(whp, cellsize = radius*2) %>% 
  st_sf(geometry = .) %>%
  filter(lengths(st_intersects(., whp)) > 0) |> 
  mutate(BASIN = 'WHP')

grd = bind_rows(ran_grd, jon_grd, whp_grd) |> 
  group_by(BASIN) |> 
  nest(grid = geometry)

# calculate the f_hab for each basin and year
f_hab = df_r |> 
  left_join(grd, by = 'BASIN') |> 
  mutate(f_sg = map2(sg_r, grid, \(r, grid) 
                     exact_extract(terra::rast(r), grid, 'mean', progress = F)[[1]]),
         f_epi = map2(lab_r, grid, \(r, grid) 
                     exact_extract(terra::rast(r), grid, 'mean', progress = F)[[1]]),
         f_man = map2(man_r, grid, \(r, grid) 
                     exact_extract(terra::rast(r), grid, 'mean', progress = F)[[1]]),
         f_alg = map2(alg_r, grid, \(r, grid) 
                     exact_extract(terra::rast(r), grid, 'mean', progress = F)[[1]])) |> 
  dplyr::select(YEAR, BASIN, grid, f_sg:f_alg) |> 
  unnest(grid:f_alg) |> 
  ungroup()|> 
  mutate(across(f_sg:f_alg, ~if_else(.x < 0, 0, .x)))

# calculate HRI for all forgaing units for each species
HRI = f_hab |>  
  slice(rep(1:n(), each = length(unique(IEI$Species)))) |> 
  mutate(Species = rep(unique(IEI$Species), times = nrow(f_hab))) |> 
  left_join(IEI, by = 'Species') |> 
  mutate(HRI = f_sg*iei_sg + f_man*iei_man + 
           f_epi*iei_epi + f_alg*iei_alg) |> 
  st_as_sf()

# save output of all E-scapes 
st_write(HRI, 'gis/FLBay_E-scapeAll.shp')

```
![*E*-scapes for a) bay anchovy, b) pinfish, and c) pink shrimp before and after the 2015 seagrass die-off for Johnson, Rankin, and Whipray basins in Florida Bay. HRI values from all years before and after were averaged for each landscape foraging unit. Warm colors (reds) represent habitat resource index (HRI) values > 1 and cool colors (blues) represent HRI values < 1. HRI values > 1 indicate greater production of resources (i.e., more energetic resources) being used by the consumer, while the opposite is true for HRI values < 1.](publication/Fig4.tiff) 

## Statistical analyses
### ANOVA
We conducted an analysis of variance (ANOVA) to compare the differences in HRI values across species, basin, and before and after the die-off. We created categorical term for if the HRI value was from before or after the seagrass die-off for the ANOVA. The years before 2015 were assigned as pre-die-off, and the years after 2015 were designated as post-die-off for our model. Species, basin, the categorical term for before or after the die-off and all interactions were used as predictor terms in the ANOVA. For significant terms in our model, pairwise comparisons using model contrasts with a Bonferroni correction were analyzed to test for significance among factor groups using the `emmeans` package. 
```{r, output=F}
library(tidyverse)
library(sf)
library(ggeffects)
library(emmeans)
library(multcomp)
library(multcompView)
```

```{r}
df = st_read('gis/FLBay_E-scapeAll.shp') |> 
  drop_na(HRI) |> 
  mutate(do = case_when(
    YEAR <= 2015 ~ 'Pre Die-off',
    YEAR > 2015 ~ 'Post Die-off'),
    do = factor(do, levels = c('Pre Die-off','Post Die-off'))) |> 
 as_tibble()


m = aov(HRI ~ do*BASIN*Species, data = df)

summary(m)

# check region comparisons
model_means = emmeans(object = m,
                      ~ do*BASIN*Species)

# add letters to each mean
model_means_cld = cld(object = model_means,
                      adjust = "bonf",
                      Letters = letters,
                      sort = F,
                      alpha = 0.05) |> 
  mutate(code = str_replace_all(.group, ' ', ''))

# show output
model_means_cld

df = left_join(df, model_means_cld, by = c('Species', 'BASIN', 'do'))

```
![Boxplot for HRI values for each species and basin pre and post 2015 seagrass die-off. Different letters indicate significantly different (p < 0.05) groups based on model contrast comparisons with a Bonferroni correction.](publication/Fig3.tiff) 

### Breakpoint 
Spatiotemporal trends in trophic function for each species were determined by averaging the HRI values of each landscape foraging unit to generate a basin-wide mean for each basin and year. We conducted breakpoint analysis across years on the mean trends on HRI value for each species using the R package `strucchange`. This package uses simple linear regression to estimate data breakpoints and test for differences in mean conditions using an intercept-only model. We assessed the optimal number of breakpoints for each model based on the lowest Bayesian information criterion (BIC) score.
```{r}
library(strucchange)

df = st_read('gis/FLBay_E-scapeAll.shp') |>
  as_tibble() |>
  group_by(Species, BASIN, YEAR) |>
  summarize(hri = mean(HRI, na.rm = T), 
            .groups = 'drop') |> 
  arrange(YEAR) |> 
  group_by(Species, BASIN) |> 
  nest() |> 
  mutate(bp = map(data, \(data) breakpoints(hri~1, data = data, h = 3)),
         fstat = map(data, \(data) sctest(Fstats(hri~1, data = data))),
         f = map_dbl(fstat, 'statistic'),
         p = map_dbl(fstat, 'p.value'),
         ci = map2(data, bp, \(data, bp) {
           if(!is.na(bp$breakpoints)) {
             confint(bp, data = data)[1]$confint
           } else {
             NA
           }})) |> 
  unnest_wider(ci, simplify = T, names_sep = '_') |> 
  mutate(change = map_dbl(data, \(data) data$YEAR[ci_1[,2]]),
         lci = map_dbl(data, \(data) data$YEAR[ci_1[,1]]),
         uci = map_dbl(data, \(data) data$YEAR[ci_1[,3]])) |> 
  dplyr::select(Species, BASIN, f, p, change, lci, uci)

df
```
![Breakpoints based on intercept model for mean habitat resource index (HRI) values for a) bay anchovy, b) pinfish, and c) pigfish for each basin. Vertical black line indicates break point and grey shade represents the 95% confidence interval. Horizontal line indicates model prediction and gray points indicate mean ± SD HRI values for basin.](publication/Fig5.tiff) 

### Correlations
We ran correlations between HRI values and $f_{habitat_i}$ across all species, years, and basins. 
```{r}
df = st_read('gis/FLBay_E-scapeAll.shp') |>
  as_tibble() 

# algae 
cor.test(x = df$f_alg, y = df$HRI, method = 'pearson')

# epipyhtes
cor.test(x = df$f_epi, y = df$HRI)

#mangrove
cor.test(x = df$f_man, y = df$HRI)

#seagrass
cor.test(x = df$f_sg, y = df$HRI,  method = 'pearson')
```

